---
title: "GEOG 5160 6160 Preprocessing examples"
author: 
  - name: "Simon Brewer"
    email: simon.brewer@ess.utah.edu
    affiliations:
      - name: University of Utah
        address: 260 S Central Campus Drive
        city: Salt Lake City
        state: UT
        postal-code: 84112
date: last-modified
format:
  html:
    toc: true
    embed-resources: true
editor: source
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(png)
library(grid)
set.seed(42)
```

# Introduction

This lab document covers a series of 'additional' topics in machine learning, mainly focused on preprocessing techniques. Although we will spend most of this semester looking at different algorithms, you will find that data manipulation makes up a significant part of any project, so some of these methods will likely be of use in your projects. 

We'll cover the following subjects:

- Feature transformation
- Encoding categorical features
- Missing data imputation
- Feature selection
- Working with imbalanced data
- Pipelines

There is no assignment for this lab - but I'd suggest at least looking over the approaches to get a sense of when these may be useful. Let's load some general libraries to get started.

```{r}
library(tidyverse)
library(tidymodels)
```

## Feature transformation

The goal of these methods is to transform individual features in some way, principally with the aim of reducing any bias in model training that results from data distributions or magnitudes.

### Simple transformation

For many methods, having continuous features that are approximately normally distributed can be important. Highly skewed data can present issues in estimating the extremes. The following are a set of simple transformation methods for both right- and left-skewed data

#### Right-skewed data

```{r message=FALSE, warning=FALSE}
## Make some right skewed data
right_skew = rnorm(1000, 0, 1)**2
hist(right_skew, col = 'darkorange')
```

```{r}
## Log transformation
right_skew_log = log(right_skew)
hist(right_skew_log)
```

```{r}
## Square root transformation
right_skew_sqrt = sqrt(right_skew)
hist(right_skew_sqrt)
```

#### Left-skewed data

```{r}
# Reflect the data around a constant (e.g., the maximum value) to make it left-skewed
left_skew = max(right_skew) - right_skew
hist(right_skew, col = 'darkorange')
```

```{r}
## Squared transformation
left_skew_sq = left_skew**2
```

```{r}
## Cube transformation
left_skew_cube = left_skew**3
```

```{r}
## Exponential transformation
left_skew_exp = exp(left_skew)
```

### Standardization and normalization

For some algorithms, having very different ranges for features (e.g. one feature in 0-1 and another in 0-1000000), can results in biases, where more attention is paid to the larger magnitude feature. These two approaches are ways to transform variables by rescaling them, with the goal of ensuring that all features have the same (or approximately) the same range. 

We'll load some example data to demonstrate this (and other steps) with the tidymodels setup. We'll use the column called `Status` as the target, and all other columns as features. Load the data and create the basic recipe to set the relationship between variables:

```{r}
data("credit_data")
```

```{r}
rec_obj <- recipe(Status ~ ., data = credit_data)
rec_obj
```

#### Standardization

Standardizing features converts them into $z$-scores by subtracting the mean and dividing by the standard deviation:
$$
z_i = (x_i - \bar{x}) / \sigma_x
$$

This has some attractive properties: it rescales all features to the same range; a value of 0 is equal to the mean; and values of -1 and 1 equal 1 s.d. below or above the mean, respectively, for all features. 

We'll now add the steps to standardize to the original recipe

```{r}
standardize <- rec_obj |> 
  step_center(all_numeric_predictors()) |> 
  step_scale(all_numeric_predictors())
```

(Note that you can combine these two steps with a function called, confusingly, `step_normalize`). To demonstrate the effect, first get the summary statistics on the original data:

```{r}
summary(credit_data)
```

Now, we do two steps: `prep` which establishes the parameters for the standardization (i.e. it calculates the mean and sd for all columns), and then `bake` to transform a dataset:

```{r}
final_rec <- prep(standardize, credit_data)
credit_data_std <- bake(final_rec, credit_data)
summary(credit_data_std)
```

If you look at the numerical variables, you should see that they are transformed to have a mean of zero and standard deviation of 1. 

#### Normalization

Normalization rescales features to a fixed range, usually $[0-1]$ (also called a min-max transformation). This is frequently used with neural networks to help in training models. We'll do the same steps as before (but with less commentary)

```{r}
normalize <- rec_obj |> 
  step_range(all_numeric_predictors()) 
```

```{r}
final_rec <- prep(normalize, credit_data)
credit_data_norm <- bake(final_rec, credit_data)
summary(credit_data_norm)
```

## Imputation

Many datasets contain missing values, usually defined by a computer as `NA` or `NaN` for 'not-a-number'. For many machine learning algorithm, these pose problems, and the functions will often fail. If your dataset contains missing values, you can either drop them (see **dplyr**s `drop_na` function, for example), or to *impute* them. This refers to replacing the missing values with a guess. We'll look at some different approaches to do that here. We'll demonstrate these here using the same dataset (you may have noticed in the `summary` output that a number of missing data are reported).

To see what this looks like, we'll print a set of rows from the data (look in the `Income` column at the 4th and 9th values).

```{r}
credit_data[150:160,]
```


We'll now try to impute these (and other missing values). The **recipes** package has a number of function, so we'll just demonstrate a few. 

### Simple imputation

We'll start with simple imputation. This replaces the missing values with a constant value, usually the mean or median. For categorical variables, we can replace with the mode or most frequent category. We'll use mean imputation, so first let's see what the mean values is for income:

```{r}
mean(credit_data$Income, na.rm = TRUE)
```

Now we'll impute. As before, we'll work from the basic recipe (`rec_obj`) that defines relationships between the variables. 

```{r}
impute <- rec_obj |> 
  step_impute_mean(all_numeric_predictors())
```

```{r}
final_rec <- prep(impute, credit_data)
credit_data_imp <- bake(final_rec, credit_data)
credit_data_imp[150:160, "Income"]
```

You should see that the previous `NA`s have been replaced with the mean (`r round(mean(credit_data$Income, na.rm=TRUE),1)`)

We can also impute categorical or nominal features using the mode. Line 30 of the data contains a missing value for `Home`:

```{r}
credit_data[30, ]
```

```{r}
impute <- rec_obj |> 
  step_impute_mode(all_nominal_predictors())
final_rec <- prep(impute, credit_data)
credit_data_imp <- bake(final_rec, credit_data)
credit_data_imp[30, ]
```

### KNN imputer

An alternative to try to make a more educated guess as to the imputed value. The KNN imputer uses a nearest neighbor approach. For a feature $x$ with a missing value for observation $i$, it searches for the closet neighboring points using all the features that are non `NaN`s. It then imputes the missing value for $x$, usually as the mean of the set of neighbors. Note that you have to set the choice of $k$, the number of neighbors that are used. This method has the advantage of working directly with both numerical and categorical data, so we can run this on all features as follows:

```{r}
impute <- rec_obj |> 
  step_impute_knn(all_predictors(), neighbors = 2)
final_rec <- prep(impute, credit_data)
credit_data_imp <- bake(final_rec, credit_data)
```

```{r}
## Numeric imputation example
credit_data_imp[150:160, "Income"]
```

```{r}
## Categorical imputation
credit_data_imp[30, ]
```

### Iterative imputer

The final method that we'll look at uses a sequence of models to impute missing data. It starts by using a simple imputation method to fill missing values (default = mean). It then iterates over each feature with missing values, creating a model of that feature as a function of all the remaining features. This is used to impute a value for each of the original missing values. It then moves on to the next feature and does the same. Once it has been through all the features, it repeats the entire process until there is only a marginal change in the imputed values, or a limit of iterations is reached. This has a large amount of flexibility: you can choose the model that is used for imputation (see the `method` argument), and it creates multiple imputation.

As there is no function to do this in the **tidymodels** library, we need to use an additional library called **mice**, so go ahead and install this now. Now let's load this:

```{r}
library(mice)
```

This comes with a neat function to visualize the pattern of missing data:

```{r}
md.pattern(credit_data)
```

Now to run the imputation, we use the `mice` function:

```{r results='hide', message=FALSE, eval=FALSE}
credit_data_imp = mice(credit_data, method = "rf")
```

You'll see a series of iterations appear on the screen. By default, this will run 5 imputations (so 5 guesses for each missing value), and it runs 5 iterations for each one to stablize the estimates). You can see the range of imputed values for the missing observations as follows:

```{r eval=FALSE}
credit_data_imp$imp$Home
```

Use the `complete` function to get the full imputed data for further analysis. The `1` refers to the imputation you want to retrieve:

```{r eval=FALSE}
credit_data_1 <- complete(credit_data_imp, 1)
credit_data_1[30, ]
```

## PCA transformation

Another useful transformation of data is through a principal component transformation. This is a commonly used form of dimension reduction, that can often reduce datasets with large numbers of features down to a small number of principal components. In addition, as PCA is based on the co-variation between features, it can avoid issues of multicollinearity or redundancy with features that record similar real world factors. We'll look at this here with a breast cancer dataset, which has a relatively large number of features (30). 

```{r}
df <- read.csv("./datafiles/breast_cancer_data.csv")
head(df)
```

Next, we'll remove some superflous columns and convert the outcome to a factor:

```{r}
df <- df |>
  select(-id, -X) |>
  mutate(diagnosis = as.factor(diagnosis))
```

We'll split into training and testing with a simple holdout

```{r}
train_test_split <- initial_split(df, prop = 0.8)

df_train <- training(train_test_split)
df_test <- testing(train_test_split)
```

Next, create a new recipe for these data:

```{r}
rec_obj <- recipe(diagnosis ~ ., data = df_train) |>
  step_normalize(all_numeric_predictors())
rec_obj
```

Now we'll build a simple initial model using logistic regression and all original, scaled features (30). We'll use the test set to evaluate and the `accuracy` function to see how well the model worked:

```{r}
## Set final recipe
final_rec <- 
  rec_obj 
## Instantiate model
log_mod <- logistic_reg()
## Build workflow linking model and recipe
log_mod_wflow <-
  workflow() |>
  add_model(log_mod) |>
  add_recipe(final_rec)
## And fit
log_mod_fit <- fit(log_mod_wflow, data = df_train)
```

Now we'll evaluate on the test set and get the accuracy:

```{r}
pred_test <- predict(log_mod_fit, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(diagnosis))

pred_test |>
  accuracy(diagnosis, .pred_class)
pred_test |>
  bal_accuracy(diagnosis, .pred_class)
```

To perform a PCA, we'll add a second stage to the recipe that includes the PCA transformation

```{r}
pca_trans <- rec_obj |> 
  step_pca(all_numeric(), num_comp = 2)
```

Now we'll rerun all the same steps as before

```{r}
## Set final recipe
final_rec <- 
  pca_trans 
## Instantiate model
log_mod <- logistic_reg()
## Build workflow linking model and recipe
log_mod_wflow <-
  workflow() |>
  add_model(log_mod) |>
  add_recipe(final_rec)
## And fit
log_mod_fit_pca <- fit(log_mod_wflow, data = df_train)
```

```{r}
pred_test <- predict(log_mod_fit_pca, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(diagnosis))

pred_test |>
  accuracy(diagnosis, .pred_class)
pred_test |>
  bal_accuracy(diagnosis, .pred_class)
```

And we see very little change in the accuracy. If you take a quick look at each fitted model (just type model name e.g. `log_mod_fit`), you will see a difference in the complexity. The first model as 30 coefficients for all the features, the second only has two, but with no loss in predictive skill. 

How do you know how many components to use? PCA can return as many summary features as original features, but of rapidly diminishing utility. A simple way to check is to independently run a PCA using R's `prcomp` function:

```{r}
df_pca <- df_train |> 
  select(-diagnosis) |> 
  prcomp(scale=TRUE)

summary(df_pca)
```

If you look at the bottom row if this output, it will show you the cumulative amount of variance as we increase the number of components. In this case, as we used the first two components, these account for about 63% of the original variance. 

## Feature selection

An alternative approach with a large number of features is to use a feature selection method. These range from simple statistical testss to more complex model-based methods. We'll look here at a couple of examples using the breast cancer data set loaded earlier. As a reminder, the original data has 30 features and 1 target:

```{r}
dim(df)
```

### Univariate feature selection

This approach does pairwise comparisons of each feature with the target using an additional step in a recipe (`step_predictor_best`). This can take a variety of methods to *score* the variables, including simple Pearson correlation, ANOVA, mutual information or random forest importance. You will need the **important** package installed to run these examples.

```{r}
library(important)
```

We'll set this up first to use the $F$-statistic from an ANOVA performed between each feature and the target. Features with low values will be excluded, and we use `prop_terms` to define the proportion of features to remove. 

```{r}
rec_obj <- recipe(diagnosis ~ ., data = df_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_predictor_best(
    all_predictors(),
    score = "aov_fstat",
    prop_terms = 0.25
  )
rec_obj
```

```{r}
df_sel <- rec_obj |>
  prep()  |> 
  bake(df) 

dim(df)
dim(df_sel)
```

We can also obtain the names of the selected/retained features:

```{r}
names(df_sel)
```

### Automated selected

**tidymodels** does not currently have an integrated function for automated selection of variables, but a number of other add-on packages do. We look here quickly at the **boruta** package (you will need to install this). ThisIn a nutshell this works by creating 'shadow' versions of each feature, then comparing importance scores from a model that includes both the original and shadow features. If the importance of any original feature is not higher than the shadow version, it is removed. The process then repeats until no more removals occur. Note that this can take a long time for large datasets due to the number iterations required.

```{r}
library(Boruta)
```

We can just run this on the original data (`df`). 

```{r}
df_bor = Boruta(diagnosis ~ ., df)
df_bor
```
For this dataset, no features were rejected, but two were deemed to be 'tentative' - running the method with more iterations may solve this. We can now get the selected features into a new set as follows (this is a bit tricky as we have keep the target column)

```{r}
new_names = names(df)[-1][df_bor$finalDecision == "Confirmed"]
df_sel = df[, c("diagnosis", new_names)]
dim(df_sel)
```

The output is a reduced data frame that can then be used in **tidymodels**.

## Imbalanced data

Next, we'll look at ways to work with imbalanced data. This generally refers to classification datasets, where the target variable has substantially more of one class than the other(s). This is very common in real world dataset, particularly anything dealing with a rare outcome. Imbalanced data can cause issues in:

- Model training: loss function is dominated by majority class
- Model evaluation: performance for minority class is under-represented

There are a range of different approaches for dealing with this, but these largely can be split into weighting methods and sampling methods. 

Next we'll load some imbalanced data (this is a data set of credit applications, with the final column indicating whether an application was rejected (`yes`) or not). We'll do some cleaning, then split into training and testing. We'll then set up an initial recipe with scaling and one-hot encoding.

```{r}
df = read.csv("./datafiles/bank-full.csv", sep=";")
head(df)
```

```{r}
df = df |>
  select(-day, -month) |>
  mutate(y = as.factor(y))

train_test_split <- initial_split(df, prop = 0.8)

df_train <- training(train_test_split)
df_test <- testing(train_test_split)
```

```{r}
rec_obj <- recipe(y ~ ., data = df_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())
```

Having got here, we can check for imbalance in the training data by calculating the imbalance ratio (number in minority class / number in majority class):

```{r}
min(table(df$y)) / max(table(df$y))
```

The imbalance ratio is around 0.13, suggsting there are nearly 10x as many `no`s and `yes`s

### Naive model

For reference, we'll fit a simple random forest model and evaluate

```{r}
## Set final recipe
final_rec <- 
  rec_obj 
## Instantiate model
rf <- rand_forest(mode='classification')
## Build workflow linking model and recipe
rf_mod_wflow <-
  workflow() |>
  add_model(rf) |>
  add_recipe(final_rec)
## And fit
rf_mod_fit <- fit(rf_mod_wflow, data = df_train)
```

```{r}
pred_test <- predict(rf_mod_fit, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(y))

pred_test |>
  accuracy(y, .pred_class)
pred_test |>
  bal_accuracy(y, .pred_class)
```

Note that although the accuracy looks great, the balanced accuracy of 0.65 reflects that one of the classes is poorly predicted.

### Resampling

In this approach, we resample the data in way that makes the number of the minority and majority class match. we need an additional library **themis** to run these functions. 

```{r}
library(themis)
```

#### Random oversampling

Here, the minority dataset is sampled *with replacement* to create a datasubset that is the same size as the majority set. To demonstrate this, we'll simply add this as an extra step, prep and bake it and count the number of `no`s and `yes`s in the returned data:


```{r}
rec_obj |>
  step_upsample(y, over_ratio = 1) |>
  prep()  |> 
  bake(new_data = NULL) |> 
  count(y, name = "training")

```

This shows the clear increase in the number of `yes`s. We'll add this an additional step when creating the final recipe, the train and evaluate a model as before:

```{r}
## Set final recipe
final_rec <- 
  rec_obj |>
  step_upsample(y, over_ratio = 1)

## Instantiate model
rf <- rand_forest(mode='classification')
## Build workflow linking model and recipe
rf_mod_wflow <-
  workflow() |>
  add_model(rf) |>
  add_recipe(final_rec)
## And fit
rf_mod_fit <- fit(rf_mod_wflow, data = df_train)
```

```{r}
pred_test <- predict(rf_mod_fit, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(y))

pred_test |>
  accuracy(y, .pred_class)
pred_test |>
  bal_accuracy(y, .pred_class)
```

#### Random undersampling

As the name implies, this downsamples the majority set to the same size as the minority.

```{r}
rec_obj |>
  step_downsample(y, under_ratio = 1) |>
  prep()  |> 
  bake(new_data = NULL) |> 
  count(y, name = "training")
```

This shows the clear decrease in the number of `no`s. We'll add this an additional step when creating the final recipe, the train and evaluate a model as before:

```{r}
## Set final recipe
final_rec <- 
  rec_obj |>
  step_downsample(y, under_ratio = 1)

## Instantiate model
rf <- rand_forest(mode='classification')
## Build workflow linking model and recipe
rf_mod_wflow <-
  workflow() |>
  add_model(rf) |>
  add_recipe(final_rec)
## And fit
rf_mod_fit <- fit(rf_mod_wflow, data = df_train)
```

```{r}
pred_test <- predict(rf_mod_fit, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(y))

pred_test |>
  accuracy(y, .pred_class)
pred_test |>
  bal_accuracy(y, .pred_class)
```

#### SMOTE

Synthetic Minority Over-sampling Technique or SMOTE. This is a different approach that creates new samples by interpolating in feature space between existing samples of the minority class. This avoids the issue of repeating samples from oversampling, and can limit some issues of overfitting

```{r}
rec_obj |>
  step_smote(y) |>
  prep()  |> 
  bake(new_data = NULL) |> 
  count(y, name = "training")
```

And added to the modeling workflow:

```{r}
## Set final recipe
final_rec <- 
  rec_obj |>
    step_smote(y)

## Instantiate model
rf <- rand_forest(mode='classification')
## Build workflow linking model and recipe
rf_mod_wflow <-
  workflow() |>
  add_model(rf) |>
  add_recipe(final_rec)
## And fit
rf_mod_fit <- fit(rf_mod_wflow, data = df_train)
```

```{r}
pred_test <- predict(rf_mod_fit, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(y))

pred_test |>
  accuracy(y, .pred_class)
pred_test |>
  bal_accuracy(y, .pred_class)
```

### Weighting

Next, we'll try to weight the dataset. One of the simplest approaches to this is to weight by the inverse of frequency (essentially $1/n_k$, where $n_k$ is the number for a given class). This is a little tricky in **tidymodels**. We need to create a new column in the data frame that contains the weight for each observation. I've set these simply here to 1 for the majority case, and 1/0.13 for the minority (to reflect the imbalance ratio. These are also set as `importance_weights` so that they are not used like other features. We'll reload and process the data to do this:

```{r}
df = read.csv("./datafiles/bank-full.csv", sep=";")

df = df |>
  select(-day, -month) |>
  mutate(y = as.factor(y),
         case_wts = ifelse(y == "no", 1, 1/0.13),
         case_wts = importance_weights(case_wts)
  )

train_test_split <- initial_split(df, prop = 0.8)

df_train <- training(train_test_split)
df_test <- testing(train_test_split)

rec_obj <- recipe(y ~ ., data = df_train) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

```

Now we'll set up everything for model training. There's one additional line here, where we specify which column contains the weights. Otherwise this is the same as before:

```{r}
## Set final recipe
final_rec <- 
  rec_obj 
## Instantiate model
rf <- rand_forest(mode = 'classification')
## Build workflow linking model and recipe
rf_mod_wflow <-
  workflow() |>
  add_model(rf) |>
  add_recipe(final_rec) |>
  add_case_weights(case_wts) ## WEIGHTS GO HERE
## And fit
rf_mod_fit <- fit(rf_mod_wflow, data = df_train)
```

And now evaluate:

```{r}
pred_test <- predict(rf_mod_fit, new_data = df_test) |>
  bind_cols(df_test |> dplyr::select(y))

pred_test |>
  accuracy(y, .pred_class)
pred_test |>
  bal_accuracy(y, .pred_class)
```

